Reduced GW invariants from cuspidal curves

### 2017-12-05 draft.tex
finished second approach, needs reviewing

Let $\mathfrak{P}=\mathfrak{Pic}^d_{\cC/\MM_1}$ with the stability condition that $\omega_\pi\otimes \mathcal L^{\otimes 3}$ is $\pi$-ample. We work over $\MM_1^{\rm{div}}$, parametrising families of nodal curves with a relative Cartier divisor. More precisely, this can be thought of as the open inside $C(\pi_*\mathcal L)=\Spec_\pP(\R^1\pi_*\mathcal L)$ (see \cite{CLpfields}), where the section is not $0$ on any irreducible component of the curve. Alternatively one can think of the moduli functor of a line bundle with a section up to scalar as the hom-stack $\Hom_{\MM_1}(\C,[\Aaff^1/\Gm])$, again requiring the section not to be $0$ and the total degree of the line bundle to be $d$.

@@

Let $E$ be the locus inside the universal curve spanned by elliptic tails; this is a Cartier divisor in the universal curve over $\MM^{wt}_1$, which we shall freely pullback and keep denoting by $E$. We write $\mathcal M:=\omega_{\pi}(E)\otimes\mathcal O_{\cC}(3D)$, where $D$ is the universal Cartier divisor over $\MM_1^{\rm{div}}$ (so $\mathcal O_{\cC}(D)$ is the pullback of the universal line bundle on $\pP$).

\begin{prop}\label{1-stabilization-div}
 The $1$-stabilisation $\MM_1^{\rm{div}}\to\MM_1^{(1),\rm{div}}$ is defined by the line bundle $\mathcal M$:
 \bcd
 (\cC,D)\ar[rr,"\phi"]\ar[dr,"\pi"] & & (\hat\cC=\operatorname{Proj}_{\MM_1^{\rm{div}}}(\bigoplus_{n\geq 0}\pi_*\mathcal M^{\otimes n}),\phi(D))\ar[ld,"\hat\pi"] \\
 & \MM_1^{\rm{div}} &
 \ecd
\end{prop}
We prove a proposition through a series of Lemma s. The first one is a technical statement telling us that we can reduce to verify semi-ampleness and local freeness on families over DVR.
\begin{lem}{Reduction to DVR}\label{DVR}
Let $T$ the spectrum of a DVR with closed point $0$ and generic point $\eta$ and let $f\colon T\to\MM_1^{\rm{div}}$ such that $0\in \mathfrak{D}^1$ and $\eta\in \MM_1^{\rm{div}}\setminus \mathfrak{D}^1.$ Then 
\[f^*\pi_*\mathcal M\cong \pi_*f^*\mathcal M.\]
We have denoted with $\mathfrak{D}^1$ the Cartier divisor image of the locus spanned by elliptic tails.
\end{lem}
\begin{proof}
The argument is taken from \cite[Lemmma~3.7.2.2]{RSPW}.
Let $K^{\bullet}$ the complex of locally free sheaves on $S$ which satisfy cohomology and base change, i.e. such that for any $T\xrightarrow{f} S$  we get
$H^i(f^*K^{\bullet})=\R^i\pi_*f^*\mathcal M.$ The construction of $K^{\bullet}$ is standard, (see e.g. \cite[Proposition~12.2]{HAR} and since the cohomology is concentrated in degree 0,1, we can assume $K^{\bullet}=K_0\to K_1.$ Let $f$ be the map defined in the statement of the Lemma.
Then we have the following exact sequences:
\begin{equation}\label{eq:exact1}
0\to \pi_*\mathcal M\to K^0\to K^1\to \R^1\pi_*\mathcal M\to 0,
\end{equation}
\begin{equation}\label{eq:exact2}
0\to \pi_*f^*\mathcal M\to f^*K^0\to f^*K^1\to \R^1\pi_*f^*\mathcal M\to 0.
\end{equation}
The derived pull-back $Lf^*$ of the complex~\eqref{eq:exact1} is zero since the complex is exact. Writing down explicitly the spectral sequence which compute the derived pull-back, we find that its convergence to zero implies that:
\begin{align*}
\operatorname{coker}(f^*\pi_*\mathcal M\to\pi_*f^*\mathcal M)&\cong \rm{L}_1f^*\R^1\pi_*\mathcal M\\
\operatorname{ker}(f^*\pi_*\mathcal M\to\pi_*f^*\mathcal M)&\cong \rm{L}_2f^*\R^1\pi_*\mathcal M.
\end{align*} 
Now, notice that $\R^1\pi_*\mathcal M$ is a line bundle supported on the Cartier divisor $\mathfrak{D}^1;$ it is clearly zero on the smooth locus and one can easily check that is rank is constant and equal to one on the divisor.
To prove that $\rm{L}_if^*\R^1\pi_*\mathcal M=0$ for $i=1,2$ it is enough to work locally on $S$, where we have a free resolution for  of the form $\R^1\pi_*\mathcal M$ of the form:
\[0\rightarrow \OO_S\xrightarrow{g}\OO_S\rightarrow \R^1\pi_*\mathcal M\rightarrow 0\] 
and thus pulling back along $f$
\[0\rightarrow  \rm{L}_1f^*\R^1\pi_*\mathcal M \OO_T\xrightarrow{f^*g}\OO_T\rightarrow f^*\R^1\pi_*\mathcal M\rightarrow 0.\] 
It immediately follows that $ \rm{L}_2f^*\R^1\pi_*\mathcal M=0.$ Finally notice that since the image of $f$ is not contained in the boundary divisor $f^*g\neq 0$ and thus injective of sheaves as $T$ is a DVR. We can then conclude $ \rm{L}_1f^*\R^1\pi_*\mathcal M=0$ as well which implies the Lemma.
\end{proof}
\begin{lem}\label{lemma:semiample}
The line bundle $\mathcal M$ is $\pi$ semi-ample, i.e. the natural map
\[\pi^*\pi_*\mathcal M^{\otimes n}\to \mathcal M^{\otimes n}\]
is surjective.
\end{lem}
\begin{proof}
Outside the locus of elliptic tails $\mathcal M$ is $\pi$- ample and thus in particular $\pi$-semi ample. To check that it is still $\pi$-semi ample also for any $x\in C$ when $C$ has an elliptic tail, thanks to the above Lemma we can reduce to check this condition when $C$ is the central fiber of a one parameter smoothing of the elliptic tail. But then this has been proved by Smyth~\cite[Lemma~2.12]{SMY1}.
\end{proof}

@@

\begin{proof}\ref{1-stabilization-div}
Let $S\to \MM_1^{\rm{div}}$ be a smooth atlas, then we have:
 \bcd
 (C,D)\ar[rr,"\phi"]\ar[dr,"\pi"] & & (\hat C=\operatorname{Proj}_{S}(\bigoplus_{n\geq 0}\pi_*\mathcal M^{\otimes n}),\phi(D))\ar[ld,"\hat\pi"] \\
 & S &
 \ecd
where: $\phi$ is proper and birational  since $\mathcal M$ is $\pi$-semi-ample and $\hat{\pi}$ is flat since $\pi_*\mathcal M$ is locally free.

Consider the caresian diagram 
\bcd
C'\ar[r,"f_C"]\ar[d,"\pi'"] & C\ar[d,"\pi"] \\
S'\ar[r,"f"] & S.
\ecd
To show that the above construction is compatible with base change it is enough to show that for any base change holds what we proved in Lemma~\ref{DVR}:
\begin{equation}\label{eq:generalbasechange}
f^*\pi_*\mathcal M\cong \pi'_*f^*\mathcal M.
\end{equation}
Since $\pi$ is flat we do have an isomorphism if we consider everything derived:
\begin{equation}\label{eq:derivedbasechange}
\rm{L}f^*\R\pi_*\mathcal M\cong \R\pi'_*\rm{L}f^*\mathcal M.
\end{equation}
First notice that the $H^0$ on the LHS of \eqref{eq:derivedbasechange} is exactly the LHS of \ref{eq:generalbasechange}.

\end{proof}

\begin{prop}
The $1$-stabilisation with divisors descends to the level of weighted curves:
\bcd
\MM_1^{\rm{div}}\ar[r]\ar[d] & \MM_1^{(1),\rm{div}}\ar[d] \\
\MM_1^{\rm{wt}}\ar[r] & \MM_1^{(1),\rm{wt}}
\ecd
\end{prop}

\subsection*{Why only cusps?} 
\subsection*{Alternative constructions}
\begin{itemize}
\item adapt the ideas of \cite{HassettHyeon}: check out section 3 of this paper;
\item check \cite[\S 4]{RSPW}.
\end{itemize}
 \section{Comparison with $p$-fields and splitting}
\subsection{Notation}
Let $\Z$ be defined by the pullback diagram:
\bcd
\mathcal Z \ar[r]\ar[d]\ar[rd,phantom,"\Box"] & \overline{\mathcal M}_1^{(1)}(\PP^4,d) \ar[d] \\
\mathcal X\ar[r] & \mathfrak M_1^{(1)}
\ecd

>>>>

Let $\mathfrak{P}=\mathfrak{Pic}^d_{\cC/\MM_1}$ with the stability condition that 
\[\omega_\pi\otimes \mathcal L^{\otimes 3}\]
 is $\pi$-ample. We work over $\MM_1^{\rm{div}}$, parametrising families of nodal curves with a relative Cartier divisor. More precisely, this can be thought of as the open inside $C(\pi_*\mathcal L)=\Spec_\pP(\R^1\pi_*\mathcal L)$ (see \cite{CLpfields}), where the section is not $0$ on any irreducible component of the curve. Alternatively one can think of the moduli functor of a line bundle with a section up to scalar as the hom-stack $\Hom_{\MM_1}(\C,[\Aaff^1/\Gm])$, again requiring the section not to be $0$ and the total degree of the line bundle to be $d$.

@@

Let $E$ be the locus inside the universal curve spanned by elliptic tails; this is a Cartier divisor in the universal curve over $\MM^{wt}_1$, which we shall freely pullback and keep denoting by $E$. We write 
\begin{equation}\label{eq:linebundlecontraction}
\mathcal N:=\omega_{\pi}(E)\otimes\mathcal O_{\cC}(3D),
\end{equation} where $D$ is the universal Cartier divisor over $\MM_1^{\rm{div}}$ (so $\mathcal O_{\cC}(D)$ is the pullback of the universal line bundle on $\pP$).
Notice that $\mathcal N$ is trivial on the locus of elliptic tails, so if the contraction morphism is well defined it will contract this locus.

\begin{prop}\label{1-stabilization-div}
 The $1$-stabilisation $\MM_1^{\rm{div}}\to\MM_1^{(1),\rm{div}}$ is defined by the line bundle $\mathcal N$:
 \bcd
 (\cC,D)\ar[rr,"\phi"]\ar[dr,"\pi"] & & (\hat\cC=\operatorname{Proj}_{\MM_1^{\rm{div}}}(\bigoplus_{n\geq 0}\pi_*\mathcal M^{\otimes n}),\phi(D))\ar[ld,"\hat\pi"] \\
 & \MM_1^{\rm{div}} &
 \ecd
\end{prop}

We shall prove this proposition through a series of lemmas. The first one is a technical tool drawn from \cite{RSPW}: it implies that the $\Proj$ construction we perform commutes with base change to generically chosen curves inside $\MM_1^{\rm{div}}$. In the following it will allow us to check that $\mathcal N$ is semi-ample and that $\pi_*\mathcal N$ is locally free assuming that $\cC\to T$ is a smoothing over the spectrum of a discrete valuation ring.

\begin{lem}[pullback with a boundary]\label{DVR}
Let $\pi\colon\cC\to S$ be a proper family of l.c.i. curves over a smooth basis, and let $\mathcal N$ be a line bundle on $\cC$ such that $\R^1\pi_*\mathcal N$ is a line bundle supported on a Cartier divisor $\Delta\subseteq S$. Then for every $T$ the spectrum of a DVR with closed point $0$ and generic point $\eta$, and for every morphism $f\colon T\to S$ such that $f(0)\in\Delta$ and $f(\eta)\in S\setminus\Delta$ we have
\[f^*\pi_*\mathcal M\cong \pi_*f^*\mathcal M.\]
%We have denoted with $\mathfrak{D}^1$ the Cartier divisor image of the locus spanned by elliptic tails.
\end{lem}
\begin{proof}
The argument can be found in \cite[Lemmma~3.7.2.2]{RSPW}.

Let $K^{\bullet}$ the complex of locally free sheaves on $S$ which satisfy cohomology and base change, i.e. such that for any $T\xrightarrow{f} S$  we get
$H^i(f^*K^{\bullet})=\R^i\pi_*f^*\mathcal M.$ The construction of $K^{\bullet}$ is standard, (see e.g. \cite[Proposition~12.2]{HAR} and since the cohomology is concentrated in degree 0,1, we can assume $K^{\bullet}=K_0\to K_1.$ Let $f$ be the map defined in the statement of the Lemma.
Then we have the following exact sequences:
\begin{equation}\label{eq:exact1}
0\to \pi_*\mathcal M\to K^0\to K^1\to \R^1\pi_*\mathcal M\to 0,
\end{equation}
\begin{equation}\label{eq:exact2}
0\to \pi_*f^*\mathcal M\to f^*K^0\to f^*K^1\to \R^1\pi_*f^*\mathcal M\to 0.
\end{equation}
The derived pull-back $Lf^*$ of the complex~\eqref{eq:exact1} is zero since the complex is exact. Writing down explicitly the spectral sequence which compute the derived pull-back, we find that its convergence to zero implies that:
\begin{align*}
\operatorname{coker}(f^*\pi_*\mathcal M\to\pi_*f^*\mathcal M)&\cong \rm{L}_1f^*\R^1\pi_*\mathcal M\\
\operatorname{ker}(f^*\pi_*\mathcal M\to\pi_*f^*\mathcal M)&\cong \rm{L}_2f^*\R^1\pi_*\mathcal M.
\end{align*} 
Now, notice that $\R^1\pi_*\mathcal M$ is a line bundle supported on the Cartier divisor $\mathfrak{D}^1;$ it is clearly zero on the smooth locus and one can easily check that is rank is constant and equal to one on the divisor.
To prove that $\rm{L}_if^*\R^1\pi_*\mathcal M=0$ for $i=1,2$ it is enough to work locally on $S$, where we have a free resolution for  of the form $\R^1\pi_*\mathcal M$ of the form:
\[0\rightarrow \OO_S\xrightarrow{g}\OO_S\rightarrow \R^1\pi_*\mathcal M\rightarrow 0\] 
and thus pulling back along $f$
\[0\rightarrow  \rm{L}_1f^*\R^1\pi_*\mathcal M \OO_T\xrightarrow{f^*g}\OO_T\rightarrow f^*\R^1\pi_*\mathcal M\rightarrow 0.\] 
It immediately follows that $ \rm{L}_2f^*\R^1\pi_*\mathcal M=0.$ Finally notice that since the image of $f$ is not contained in the boundary divisor $f^*g\neq 0$ and thus injective of sheaves as $T$ is a DVR. We can then conclude $ \rm{L}_1f^*\R^1\pi_*\mathcal M=0$ as well which implies the Lemma.
\end{proof}
\begin{lem}\label{lemma:semiample}
The line bundle $\mathcal M$ is $\pi$ semi-ample, i.e. the natural map
\[\pi^*\pi_*\mathcal M^{\otimes n}\to \mathcal M^{\otimes n}\]
is surjective.
\end{lem}
\begin{proof}
Outside the locus of elliptic tails $\mathcal M$ is $\pi$- ample and thus in particular $\pi$-semi ample. To check that it is still $\pi$-semi ample also for any $x\in C$ when $C$ has an elliptic tail, thanks to the above Lemma we can reduce to check this condition when $C$ is the central fiber of a one parameter smoothing of the elliptic tail. But then this has been proved by Smyth~\cite[Lemma~2.12]{SMY1}.
\end{proof}

@@

\begin{proof}\ref{1-stabilization-div}
Let $S\to \MM_1^{\rm{div}}$ be a smooth atlas, then we have:
 \bcd
 (C,D)\ar[rr,"\phi"]\ar[dr,"\pi"] & & (\hat C=\operatorname{Proj}_{S}(\bigoplus_{n\geq 0}\pi_*\mathcal M^{\otimes n}),\phi(D))\ar[ld,"\hat\pi"] \\
 & S &
 \ecd
where: $\phi$ is proper and birational  since $\mathcal M$ is $\pi$-semi-ample and $\hat{\pi}$ is flat since $\pi_*\mathcal M$ is locally free. 
To verify that this define a morphism $S\to \MM_1^{\rm{div}}(1)$ we have to argue that $\hat{C}$ has reduced fibers and only nodes and cusps as singularities.

Since these properties only concern the fiber of $\hat{\pi}$ so we can verify them after base change to a DVR $T$ and we can choose it as in Lemma~\ref{DVR} so that the construction commutes with base change.

But then this is just the statement of Smyth's contraction Lemma~\cite[Lemma~2.13]{SMY1} if we assume $C_T$ smooth and, otherwise, one can argue following \cite[Proposition~3.7.3.1]{RSPW}:
being $C_T$ a smoothing of a nodal curve over a DVR it is normal by Serre's criterion of normality (($R_1$), ($S_2$)). Moreover $C_0\setminus E\to\hat{C}_0$ is birational, which implies $\hat{C}_0$ is reduced as well. 
Finally, the fact that $\hat{C}_0$ can only acquire a cusp as singularity simply follows from the classification of genus 1 uni-branch singularities \cite[Appendix~A]{SMY1}

Finally, to conclude that this define a morphism 
\[\MM_1^{\rm{div}}\to \MM_1^{\rm{div}}(1)\]
it is enough to verify that
 $\rm{pr}_1(\phi)\cong\rm{pr}_2(\phi)$ where $\rm{pr}_i\colon S'= S\times_{\MM_1^{\rm{div}}} S\to S$ and that $\rm{pr}_{ij}\colon S\times_{\MM_1^{\rm{div}}} S\times_{\MM_1^{\rm{div}}} S\to  S\times_{\MM_1^{\rm{div}}} S $ satisfy the cocycle condition.
 The first condition follows from the fact that $\rm{pr}^*_i\phi$ are the morphism defined by 
 $\rm{pr}_i^*\pi_{s,*}\mathcal{M}\cong\pi_ {S',*} \rm{pr}_i^*\mathcal M.$ Thus it is enough to show that 
 \[\rm{pr}_1^*\mathcal M\cong \rm{pr}_2^*\mathcal M.\]
  But $\mathcal M$ is actually pulled back from $\MM_1^{\rm{div}}$ and thus the wanted isomorphism just follows from the commutativity of the following diagram

 \bcd
S\times_{\MM_1^{\rm{div}}} S\ar[r,"\rm{pr_1}"]\ar[d, "\rm{pr_2}"] & S\ar[d] \\
S\ar[r] & \MM_1^{\rm{div}}
\ecd

The  cocycle condition is similarly straightforward to verify. 
\end{proof}

\begin{prop}
The $1$-stabilisation with divisors descends to the level of weighted curves:
\bcd
\MM_1^{\rm{div}}\ar[r]\ar[d] & \MM_1^{(1),\rm{div}}\ar[d] \\
\MM_1^{\rm{wt}}\ar[r] & \MM_1^{(1),\rm{wt}}
\ecd
\end{prop}
\begin{proof}
{\'E}tale locally on $\MM_1^{\rm{wt}}$ we can always choose sections $s_i$s such that the associated Cartier divisor $D$ satisfies that $\omega_{\pi}(E)\otimes\mathcal O_{\cC}(3D)$ is trivial on elliptic tails and ample else-where.
Then if $S\to \MM_1^{\rm{wt}}$ is a smooth atlas, up to take an {\'e}tale cover we can assume there are sections  $s_i$ s of $\mathcal C_S\xrightarrow{\pi_S} S$ defining a lifting $S\to \MM_1^{\rm{div}}$ and thus trough the above construction a morphism $S\to  \MM_1^{(1),\rm{wt}}.$

Now to show this define a morphism $\MM_1^{\rm{wt}}\rightarrow \MM_1^{(1),\rm{wt}}$ we need once again to verify that, in the notation of the above Proposition,  $\rm{pr}_1(\phi)\cong\rm{pr}_2(\phi)$  and the cocycle condition is satisfied.

In other words, we need to check that for two different lifting $S\xrightarrow{s_i,\;s'_i}\MM_1^{\rm{div}}$, 
\[\hat{\mathcal C}_1=\operatorname{Proj}_S\left((\omega_{\pi}(E)\otimes\mathcal O_{\cC}(D_1))^{\otimes n}\right)\cong \operatorname{Proj}_S\left((\omega_{\pi}(E)\otimes\mathcal O_{\cC}(D_2))^{\otimes n}\right)=\hat{\mathcal C}_2. \]
By construction, we have a birational map 
 \bcd
&\mathcal C \ar[dl,"\phi_1"]\ar[dr,"\phi_2"]  \\
 \hat{\mathcal C}_1\ar[rr,dashed, "\psi"] & &\hat{\mathcal C}_2.
 \ecd

We want to show that it extends to a regular morphism. 
First of all, notice that $ \hat{\mathcal C}_i$ are normal. Indeed, since $S$ is smooth and the singularities of the fibers are in co-dimension 1, both $ \hat{\mathcal C}_i$ are regular in codimension 1. Moreover, since both $S$ (smooth) and the fibers (Cohen-Macaulay) satisfy the ($S_2$) condition of Serre, also the total spaces do\cite[Thorem~23.9]{MAT}. In particular, this implies  that
\[\phi_{i,*}\OO_{\mathcal C}\cong \OO_{\hat{\mathcal C}_i} \]
 Moreover notice that by construction $\operatorname{Ex}(\phi_1)\cong \operatorname{Ex}(\phi_2),$ so in particular $\phi_2$ contracts all fibers of $\phi_1.$ Then a standard argument \cite{debarre}
shows that $\phi_2$ factorise through $\phi_1$ and vice versa. This prove the existence of the isomorphism $\psi.$ Notice that this is in fact unique as it is the one making the diagram commutative.

Likewise we can show the cocycle condition and thus the proof is completed.
\end{proof}

\subsection*{Why only cusps?}

 \section{Comparison with $p$-fields and splitting}
\subsection{Notation}
Let $\Z$ be defined by the pullback diagram:
\bcd
\mathcal Z \ar[r]\ar[d]\ar[rd,phantom,"\Box"] & \overline{\mathcal M}_1^{(1)}(\PP^4,d) \ar[d] \\
\mathcal X\ar[r] & \mathfrak M_1^{(1)}
\ecd
